The Ex-Ante View of Recommender System Design

### 2019-09-17 paper/filter_bubbles.tex
updates to paper

\usepackage{xspace}
\newcommand{\OMIT}[1]{}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations. While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.

\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making in the context of large choice sets faced by users in online markets. We then ask how a stylized model of recommendation can shape the behavior of long-lived users and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.

We focus on understanding the degree to which recommender systems can lead users into \textit{filter bubbles}. Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}. The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated. This means that when users consume a good and learn its true utility this gives them information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty. This learning spill-over induces users to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns. This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.

We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns. We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences. We model existing recommendation systems as providing users with information on the common-value component and denote as this as partial recommendation. We consider how user behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for users if they knew the true utility of all the goods in their choice set.

Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst users.

@@

Stacking utilities in vector-form, we get:
\begin{align*}
{\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V
\end{align*}
We stack the idiosyncratic and common-value components as $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
\par
\noindent \textbf{user Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.  Formally, user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal. The idiosyncratic utility and common-value utility are distributed as follows:
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}

We impose the normality assumption for two reasons. The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.

>>>>

\usepackage{color-edits}
\addauthor{ss}{red} 
\usepackage{xspace}
\newcommand{\OMIT}[1]{}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{https://www.mckinsey.com/industries/retail/our- insights/how-retailers-can-keep-up-with-consumers} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.

\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making of users in online markets \textit{without recommendation} and investigate whether the negative effects associated with recommender systems arise even in the absence of recommendation. We then ask how a stylized model of recommendation can shape the behavior of long-lived users and under what conditions it amplifies or deflates these effects. Finally, we use the insights from our model to improve the design and evaluation of recommender systems.

The key components of our model are:
\begin{enumerate}
\item Users have uncertainty about their preferences over goods and thus face a decision problem under uncertainty.
\item Users face large choice sets and only consume a small portion of this choice set over their lifetime.
\item Users are long-lived and sequentially consume items. Consumption of an item has learning spill-overs where consumption gives information about the utilities of similar items in the product space.
\item Recommendation provides users with recommendation to reduce their uncertainty.
\end{enumerate}
Using this model we first investigate the degree to which recommender systems can lead users into filter bubbles. Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation. The key component of our model that drives this result is that the utilities of similar goods are correlated and that, when users consume a good, it gives them information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty. This learning spill-over induces users to consume goods ``similar" to those they consumed before that had high realized utility leading to an increasing narrowing of consumption towards these regions of the product space. This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty. We consider recommendation as providing information to consumers in order to reduce their uncertainty and we find that this induces them to explore a more diverse set of products, which is consistent with the empirical results from \cite{nguyen2014exploring}.

We further explore how diverse the overall set of items that users consume is and find that, without recommendation, the diversity of items consumed is negatively correlated with welfare. This is due to the fact that, once users find good items, they consume items nearby in the product space due to the correlation of utilities. When they consume bad products, they move far away in the product space leading to a more diverse set of consumed products but lower welfare. However, when users are very risk-averse we find that they may stay in small portions of the product space. Thus, we find that consumption set diversity is not by a itself a useful measure for understanding the impact of recommender systems on users.

We find that recommendation can lead to increases in homogeneity amongst users and that, without recommendation, users are almost fully inhomogenous. We model the realized utilities of users as being a weighted sum of a common-value and idiosyncratic component. The recommender can only learn the common-value component of utility since the idiosyncratic component is inherently unpredictable given other individuals' preferences. The recommender uses its knowledge of the common-value component and individuals' beliefs over the product space when designing personalized recommendation. This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component which leads to an increase in homogeneity.

Finally, we apply our model and findings to inform the design and evaluation of recommender systems. Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation. Thus, in order to evaluate the value of recommendation it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior. This requires collecting data that is not traditionally collected by recommender systems - the beliefs that users have of the utilities of goods and understanding the degree of uncertainty users have over different portions of the product space.

@@

Stacking utilities in vector-form, we get:
\begin{align*}
{\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V
\end{align*}
We stack the idiosyncratic and common-value components as $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
\par
\noindent \textbf{user Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.  Formally, user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal. The idiosyncratic utility and common-value utility are distributed as follows:
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with \ssedit{$\overline V =0$}
\end{enumerate}

We impose the normality assumption for two reasons. \ssedit{The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.} The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
